Propositions as Sessions, Mechanically

### 2015-04-18 paper/conclusion.tex
Spell checking and minor tweaks.

The current paper presented work towards a formalisation of
\citeauthor{Wadler:2014}'s two-tier GV/CP system. The Coq development defines
the process calculus CP and the functional language GV using state-of-the-art
techniques from the programming language and interactivee theorem proving
communities. Proof of subject reduction for CP has been formalised along with
work towards proving sound both cut elimination and the GV to CP CPS
translation. While ultimately there remains work to be done to complete the
development, several issues and guidelines have been identified to be
considered by future endeavours in the area. Finally, some ideas for future
work, on GV/CP specifically but also generally in the area of session-based
type system formalisations, were proffered.

>>>>

The current paper presented work towards a formalisation of
\citeauthor{Wadler:2014}'s two-tier GV/CP system. The Coq development defines
the process calculus CP and the functional language GV using state-of-the-art
techniques from the programming language and interactive theorem proving
communities. Proof of subject reduction for CP has been formalised along with
work towards formalising both cut elimination and the GV to CP CPS
translation. While ultimately there remains work to be done to complete the
development, several issues and guidelines have been identified to be
considered by future endeavours in the area. Finally, some ideas for future
work on GV/CP specifically but also generally in the area of session-based
type system formalisations were proffered.

### 2015-04-18 paper/formalisation.tex
Spell checking and minor tweaks.

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a $\stlc$ this is not onerous, but GV has four
binding constructs, CP has six and propositions have two. Indeed, a lot of
effort was expended setting up this infrastructure. However, I believe one of
the contributions of this work is to articulate what one needs from state of
the art metatheory libraries in order to handle session-based languages, and
more generally, linear type systems.

@@

Ordering in environments is ignored so the development required operations for
handling permutations of environments as shown in the \coqe$cp_rule$
definition. The Metatheory library does not currently support permutating
environments, however some basic lemmata were added as a result of this work,
and custom tactics were developed to automatically solve certain permutation
goals via simple proof search. Even so, there are still many cases where
sequences of primitive lemmata, such as those handling append, commutativity
and transitivity, need to be applied manually. A lot of these cases are
similar but it would not be advantageous to develop a tactic to do the
work. Doing so would create a dependency between a hard-to-debug and obscure
Ltac definition and the proofs which utilise it, becoming a burden to
change. Further, such a development philosophy treats Ltac tactics akin to
macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
is doubtful much can be gained from it unless the execution of Ltac tactics
can be made more transparent.

@@

\coqe$Theorem$ or \coqe$Lemma$ indicate the start of a proposition to be
proved (Coq makes no distinction between the two keywords). The statement to
prove is that if \coqe$P$ is a well-typed process in a context $\Gamma$ and
\coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is well-typed in $\Gamma$. The
proof environment is started using \coqe$Proof.$ and then followed by a
sequence of tactics until all goals have been proven. After execution of
\coqe$Qed.$ the Coq system inspects the validity of the proof object
constructed by the tactics and, if valid, will add the theorem to the
environment.

Some tactics used above are shorthands for commonly combined tactics, for
example \coqe$ii$ repeatedly performs hypothesis introduction then
simplification until no more can be peformed, and \coqe$gen$ $\Gamma$
generalises the goal over the typing environment $\Gamma$.

@@

A consequence of defining the reduction relation with respect to compound
process terms is that the subprocesses have implicit derivations which are
unavailable in the conclusion. For example, \coqe$reduce_spawn$ provides no
information about the environment which types \coqe$P$ in the statement of the
lemma. The derivation for \coqe$P$ (similarly, \coqe$Q$) is obtained by
inverting the hypothesis \key{WT} and specialising the resulting assumptions
to a newly introduced fresh variable (to replace the de Bruijn index). In the
case of \coqe$reduce_spawn$, two inversions are required to get derivations
for \coqe$P$ and \coqe$Q$. This property of the reduction relation affects the
definitions of the typing rules for CP, though it was not obvious at the time.
Notable examples are the structural rules, weakening and contraction, and
empty choice.

\textbf{Weakening and Contraction.} Initially, an implicit notion of weakening
was chosen so as to be in line with \citeauthor{Wadler:2014}'s
definitions. However, explicit weakening is required to differentiate the
principal cut elimination rules for the interactions between servers and
clients; spawning and garbage collection of a server with no clients
(weakening). The paper presentation does not specify where weakening could
occur due to its implicit nature and it is tacitly assumed weakening can
always be pushed further up the derivation tree to a base rule
e.g. \coqe$cp_fwd$ rule. The same is true of contraction where it is defined
as a substitution of names on process terms, not a syntax-directed rule.

@@

The reader may wonder how weakening is supported if not
contraction. Fortunately, weakening only requires the names (not the types in
the bindings) within the environment, and these may be obtained by relying on
the following lemma relating the free variables of a process and the domain of
its environment:
\begin{coq}
Lemma in_env_fv:
  forall P $\Gamma$ x (WT: P $\tpcp$ $\Gamma$),
    x $\in$ dom $\Gamma$ $\iff$ x $\in$ fv_proc P.
\end{coq}

@@

where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
rule for \coqe$cp_cut$ does not specify any particular split of the linear
environment in the conclusion; splitting is non-deterministic. So, in order to
prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
on \coqe$open_proc P x$. Similarly, to prove the converse statement
\coqe$forall x, open_proc R x$ must be assumed. A different associativity rule
would occur if \coqe!Q! was locally closed in the above. For commuting
conversions, permutating binders is essential in formulating the majority of
the rules.

@@

After inverting \key{WT} and specialising the resulting hypotheses regarding
the subprocesses with fresh variables the proof state is as follows:

@@

Following informal proofs, it seems likely that the proof requires a
termination measure involving the length of the derivation. In hindsight
however, I do not believe the current encoding is the best for this sort of
reasoning. Formalised efforts for proving cut elimination for other logics
have adopted an approach which permits reasoning on the structure of
derivations represented explicitly as
trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
here, that information is limited and ephemeral in that it is only available
during the proving of a derivation e.g. a reduction lemma. It is future work
to investigate these and other strategies which will assist in formalising
cut-elimination for CP.

@@

The typing judgements for the functional language GV are shown in
Figure~\ref{fig:gv}. Note that I have altered the $n$-ary branch and choice
constructs to the restricted binary versions in order to avoid tricky
formulations. The change is not restrictive however, since CP itself provides
only binary versions of plus ($\oplus$) and with ($\with$)
constructs. Explicit limiting and unlimiting of abstraction was chosen to
enable the inclusion of these rules in the translation from GV to CP, which is
performed by induction on GV terms and is therefore
syntax-directed. Figure~\ref{fig:letscompare} presents the informal version of
the well-typed let rule. While \citeauthor{Char:2012} suggests using
two-indices for handling multiple binders~\cite{Char:2012}, such a strategy
seems excessive for the GV let rule and the only rule to benefit from multiple
binders. For that reason, the simple solution of performing two openings was
chosen.

@@

Another issue related to types is how to represent type kinds, linear and
unlimited kind. Initially the type definition was indexed by the kind and
while this definition has the advantage of preventing certain ill-formed
types, such as a non-linear session type, it complicated later definitions and
proofs. For instance, the decidability of equality on types now relied on an
axiom permitting heterogeneous equality~\cite{McBride:1999}. Generally, it is
better to avoid unnecessary postulates if possible. Further, when formulating
the typing rules, the kind must be given to each type hypothesised and most
rules are polymorphic in the kind, requring an awkward kind inspection
function to be applied to each type within a binding. The alternative approach
taken was to define a well-formed type relation and remove the kind index from
the type definition. Figure~\ref{fig:wft} shows the well-formedness for the
linear function type. Notice, that even though the conclusion must have linear
kind, as in rule \coqe!wt_tm_labs!, the argument and result types can have any
kind, highlighting the aforementioned kind polymorphism. An additional
predicate was defined to ensure typing rules consider only valid session
types. The predicate appears in the well-formedness constructors for session
types, ensuring the continuation is a session type and not, for example, a
linear function type. These issues highlight that one should be mindful of the
chosen representation and its effect on later expressiveness.

@@

The translation of from GV types to CP propositions is shown in
Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
taking a GV session type to its notional dual in CP, reflecting the semantic
difference between channels in GV and those in CP. For GV, the type refers to
the action on the channel endpoint. For CP, the proposition refers to the
\textit{interface} exposed to a user of the channel. For example, output in GV
corresponds to input in CP; the action on the channel endpoint is to output a
value, and to do this the channel must input the value provided by the
programmer.

@@

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, and then define a dependent function that only
considers the positive cases for which the relation is
defined~\cite{McKinna:2009}. An example, translating choice, is shown in
Figure~\ref{fig:trans}. The translation of choice considers \coqe!NL! and
\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
term in the combined environment. The \coqe!NL! and \coqe!NR! subterms need to
be considered as opened terms using cofinite quantification in order to yield
suitable induction hypotheses. The graph relation is able to specify these
constraints in contrast to the monadic function definition. However, several
issues were encountered when attempting to specify the translation. The graph
relation requires much of the clutter present in both the \coqe!cp_rule! and
\coqe!wt_tm! definitions such as linear enviornments and cofinite
quantification, and unfortunately the specification of choice in
Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! is not being
opened in the \key{TR} assumption yet in the conclusion it is not locally
closed. It would appear that, in addition to opening GV subterms, the graph
relation must also open CP subprocesses even in cases where there was no
binder in the original GV term e.g. function
application~\cite{Wadler:2014}. Attempts to resolve this issue are left for
future work as is the proof of \coqe!cps_trans_wt!, but it appears the graph
relation approach is the most promising. These difficulties again illustrate
the importance of the representation choice.

>>>>

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a $\stlc$ this is not onerous, but GV has four
binding constructs, CP has six and propositions have two. Indeed, a lot of
effort was expended setting up this infrastructure. However, I believe one of
the contributions of this work is to articulate what one needs from
state-of-the-art metatheory libraries in order to handle session-based
languages, and more generally, linear type systems.

@@

Ordering in environments is ignored so the development required operations for
handling permutations of environments as shown in the \coqe$cp_rule$
definition. The Metatheory library does not currently support permuting
environments, however some basic lemmata were added as a result of this work,
and custom tactics were developed to automatically solve certain permutation
goals via simple proof search. Even so, there are still many cases where
sequences of primitive lemmata, such as those handling append, commutativity
and transitivity, need to be applied manually. A lot of these cases are
similar but it would not be advantageous to develop a tactic to do the
work. Doing so would create a dependency between a hard-to-debug and obscure
Ltac definition and the proofs which utilise it, becoming a burden to
change. Further, such a development philosophy treats Ltac tactics akin to
macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
is doubtful much can be gained from it unless the execution of Ltac tactics
can be made more transparent.

@@

\coqe$Theorem$ or \coqe$Lemma$ indicate the start of a proposition to be
proved (Coq makes no distinction between the two keywords). The statement to
prove is that if \coqe$P$ is a well-typed process in a context $\Gamma$ and
\coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is well-typed in $\Gamma$. The
proof environment is started using \coqe$Proof.$ then followed by a sequence
of tactics until all goals have been proven. After execution of \coqe$Qed.$
the Coq system inspects the validity of the proof object constructed by the
tactics and, if valid, will add the theorem to the environment.

Some tactics used above are shorthands for commonly combined tactics, for
example \coqe$ii$ repeatedly performs hypothesis introduction then
simplification until no more can be performed, and \coqe$gen$ $\Gamma$
generalises the goal over the typing environment $\Gamma$.

@@

A consequence of defining the reduction relation with respect to compound
process terms is that the subprocesses have implicit derivations which are
unavailable in the conclusion. For example, \coqe$reduce_spawn$ provides no
information about the environment which types \coqe$P$ in the statement of the
lemma. The derivation for \coqe$P$ (similarly, \coqe$Q$) is obtained by
inverting the hypothesis \key{WT} and specialising the resulting assumptions
to a newly introduced fresh variable (replacing the de Bruijn index). In the
case of \coqe$reduce_spawn$, two inversions are required to get derivations
for \coqe$P$ and \coqe$Q$. This property of the reduction relation affects the
definitions of the typing rules for CP, though it was not obvious at the time.
Notable examples are the structural rules, weakening and contraction, and
empty choice.

\textbf{Weakening and Contraction.} Initially, an implicit notion of weakening
was chosen so as to be in line with \citeauthor{Wadler:2014}'s
definitions. However, explicit weakening is required to differentiate the
principal cut elimination rules for the interactions between servers and
clients: spawning and garbage collection of a server with no clients
(weakening). The paper presentation does not specify where weakening could
occur due to its implicit nature and it is tacitly assumed weakening can
always be pushed further up the derivation tree to a base rule
e.g. \coqe$cp_fwd$ rule. The same is true of contraction where it is defined
as a substitution of names on process terms, not a syntax-directed rule.

@@

The reader may wonder how weakening is supported if not
contraction. Fortunately, weakening only requires the names (not the types in
the bindings) within the environment, and these may be obtained by relying on
the following lemma relating the free variables of a process and the domain of
its environment:\pagebreak
\begin{coq}
Lemma in_env_fv:
  forall P $\Gamma$ x (WT: P $\tpcp$ $\Gamma$),
    x $\in$ dom $\Gamma$ $\iff$ x $\in$ fv_proc P.
\end{coq}

@@

where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
rule for \coqe$cp_cut$ does not specify any particular split of the linear
environment in the conclusion; splitting is non-deterministic. So, in order to
prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
on \coqe$open_proc P x$. Similarly, to prove the converse statement
\coqe$forall x, open_proc R x$ must be assumed. A different associativity rule
would occur if \coqe!Q! was locally closed in the above. For commuting
conversions, permuting binders is essential in formulating the majority of
the rules.

@@

After inverting \key{WT} and specialising the resulting hypotheses regarding
the subprocesses to fresh variables the proof state is as follows:

@@

Following informal proofs, it seems likely that the proof requires a
termination measure involving the length of the derivation. In hindsight
however, I do not believe the current encoding is the best for this sort of
reasoning. Formalised efforts for proving cut elimination for logics, rather
than process calculi, have adopted an approach which permits reasoning on the
structure of derivations represented explicitly as
trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
here, that information is limited and ephemeral in that it is only available
during the proving of a derivation e.g. a reduction lemma. It is future work
to investigate these and other strategies for formalising cut-elimination for
CP.

@@

Some typing judgements for the functional language GV are shown in
Figure~\ref{fig:gv}. Note that I have altered the $n$-ary branch and choice
constructs to the restricted binary versions in order to avoid tricky
formulations. The change is not restrictive however, since CP itself provides
only binary versions of plus ($\oplus$) and with ($\with$)
constructs. Explicit limiting and unlimiting of abstraction was chosen to
enable the inclusion of these rules in the translation from GV to CP, which is
performed by induction on GV terms and is therefore
syntax-directed. Figure~\ref{fig:letscompare} presents the informal version of
the well-typed let rule. While \citeauthor{Char:2012} suggests using
two-indices for handling multiple binders~\cite{Char:2012}, such a strategy
seems excessive for the GV let rule and the only rule to benefit from multiple
binders. For that reason, the simple solution of performing two openings was
chosen.

@@

Another issue related to types is how to represent type kinds, linear and
unlimited kind. Initially the type definition was indexed by the kind and
while this definition has the advantage of preventing certain ill-formed
types, such as a non-linear session type, it complicated later definitions and
proofs. For instance, the decidability of equality on types now relied on an
axiom for heterogeneous equality~\cite{McBride:1999}. Generally, it is better
to avoid unnecessary postulates if possible. Further, when formulating the
typing rules, the kind had to be given to each type hypothesised and most
rules are polymorphic in the kind, requiring an awkward kind inspection
function to be applied to each type within a binding. The alternative approach
taken was to define a well-formed type relation and remove the kind index from
the type definition. Figure~\ref{fig:wft} shows the well-formedness for the
linear function type. Notice, that even though the conclusion must have linear
kind, as in rule \coqe!wt_tm_labs!, the argument and result types can have any
kind, highlighting the aforementioned kind polymorphism. An additional
predicate was defined to ensure typing rules consider only valid session
types. The predicate appears in the well-formedness constructors for session
types, ensuring the continuation is a session type and not, for example, a
linear function type. These issues highlight that one should be mindful of the
chosen representation and its effect on later expressiveness.

@@

The translation from GV types to CP propositions is shown in
Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
taking a GV session type to its notional dual in CP, reflecting the semantic
difference between channels in GV and those in CP. For GV, the type refers to
the action on the channel endpoint. For CP, the proposition refers to the
\textit{interface} exposed to a user of the channel. For example, output in GV
corresponds to input in CP; the action on the channel endpoint is to output a
value, and to do this the channel must input the value provided by the
programmer.

@@

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, then define a dependent function that only considers
the positive cases for which the relation is defined~\cite{McKinna:2009}. An
example, translating choice, is shown in Figure~\ref{fig:trans}. The
translation of choice considers \coqe!NL! and \coqe!NR! in a separate
environment to \coqe!M! and types the resulting case term in the combined
environment. The \coqe!NL! and \coqe!NR! subterms need to be considered as
opened terms using cofinite quantification in order to yield suitable
induction hypotheses. The graph relation is able to specify these constraints
in contrast to the monadic function definition. However, several issues were
encountered when attempting to specify the translation. The graph relation
requires much of the clutter present in both the \coqe!cp_rule! and
\coqe!wt_tm! definitions such as linear environments and cofinite
quantification, and unfortunately the specification of choice in
Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! in the
conclusion is not locally closed yet it is not being opened in the \key{TR}
assumption. It would appear that, in addition to opening GV subterms, the
graph relation must also open CP subprocesses even in cases where there was no
binder in the original GV term e.g. function
application~\cite{Wadler:2014}. Attempts to resolve this issue are left for
future work as is the proof of \coqe!cps_trans_wt!, but it appears the graph
relation approach is the most promising. These difficulties again illustrate
the importance of the representation choice.

### 2015-04-18 paper/futurework.tex
Spell checking and minor tweaks.

Further progress with the graph relation version of the translation could be
made in the current setting aiming to limit the duplication discussed
previously. The translation would be substantially affected by the
aforementioned change to CP because the derivations in the conlusions of the
rules would have to be constructed. In this case, it may be worth
investigating representing derivations in GV explicitly.

@@

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources. The ``adopt'' permits aliases on linear resources and ``focus'' is
used to access a linear resource through one of the aliases. The type system
requires special language features to manage
capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
or soundness results for the type system. It would be interesting to
investigate these properties formally in a session type setting. The ``focus''
rule enforces an invariant on the focussed object (channel) such that the type
is maintained during the focus operation. The object may be aliased while it
is under a focus operation. Such a rule would appear to only benefit recursive
session types. Indeed, focussing on a non-recursive channel would appear
pointless because one is prevented from communicating across it for the entire
scope of the focus. Moreover, it would be undesirable to permit modification
of a session type without guaranteeing exclusive access to the channel since
this could lead to invalid communication.

>>>>

Further progress with the graph relation version of the translation could be
made in the current setting aiming to limit the duplication discussed
previously. The translation would be substantially affected by the
aforementioned change to CP because the derivations in the conclusions of the
rules would have to be constructed. In this case, it may be worth
investigating representing derivations in GV explicitly.

@@

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources. The ``adopt'' permits aliases on linear resources and ``focus'' is
used to access a linear resource through one of the aliases. The type system
requires special language features to manage
capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
or soundness results for the type system. It would be interesting to
investigate these properties formally in a session type setting. The ``focus''
rule enforces an invariant on the focussed object (channel) such that the type
is maintained during the focus operation. Such a rule would appear to only
benefit recursive session types. Indeed, focussing on a non-recursive channel
would appear pointless because one is prevented from communicating across it
for the entire scope of the focus. Moreover, it would be undesirable to permit
modification of a session type without guaranteeing exclusive access to the
channel since this could lead to invalid communication.
